---
title: "<Species> SRP Pipeline Report"
author: "Tristan Sones-Dykes"
date: "31/07/23"
output: html_document
---

```{r setup, include=FALSE}
library(here)
# grab functions from src
source(here("src", "signal_peptides.r"))
source(here("src", "hydrophobicity.r"))

# define paths
protein_path <- here("data", "Proteins", "tslil_protein_strings.fasta")
RNA_path <- here("data", "RNA", "SC_999.fasta")

# read in AA and RNA strings
AA_stringset <- readAAStringSet(protein_path)
RNA_stringset <- readDNAStringSet(RNA_path)

# run signalP for protein names
protein_names <- signalp(protein_path)

# run phobius on signalP results
outfile <- tempfile(fileext = ".fasta")
writeXStringSet(AA_stringset[protein_names], outfile)
TM_df <- phobius(outfile)
```

### Summary
In this doc we use SignalP to find signal peptides in AA strings, then run phobius
and use the Kyte-Doolittle hydrophobicity scale to find a compound hydrophobicity score
to separate secreted proteins into ones that use the signal recognition particle (SRP) and 
those that don't. Then use motif analysis to find out if Ssd-1 significantly targets one 
of these groups.

## SignalP and Hydrophobicity 
After running SignalP and Phobius, we can use the Kyte-Doolittle hydrophobicity scale to
find the compound hydrophobicity score for each protein with a signal peptide, then find
the underlying bimodal distribution to categorise the proteins into the two SRP groups.

```{r}
library(mixtools)

# find hydrophobicity scores for each protein
compound_df <- data.frame(geneID = character(), compound_hydropathy_score = numeric())
for(i in seq_len(nrow(TM_df))) {
    hydrophobicity_df <- hydropathy_local(AA_stringset[TM_df$ID[i]])
    max_hydrophobicity <- hydrophobicity_df %>%
        filter(Position >= TM_df$start[i]) %>%
        filter(Position <= TM_df$end[i]) %>%
        summarise(max_hydrophobicity = max(WindowHydropathy)) %>%
        pull(max_hydrophobicity)

    window_length <- TM_df$end[i] - TM_df$start[i] + 1
    compound_hydropathy_score <- max_hydrophobicity * window_length
    compound_df <- rbind(compound_df, data.frame(geneID = TM_df$ID[i], compound_hydropathy_score = compound_hydropathy_score))
}

mixed_model <- normalmixEM(compound_df$compound_hydropathy_score, k = 2)

# plot distribution of scores with mixture model using ggplot2
# with an overlay of the two normal distributions
ggplot(compound_df, aes(x = compound_hydropathy_score)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "white", colour = "black") + 
    stat_function(fun = dnorm, args = list(mean = mixed_model$mu[1], sd = mixed_model$sigma[1]), colour = "red") +
    stat_function(fun = dnorm, args = list(mean = mixed_model$mu[2], sd = mixed_model$sigma[2]), colour = "green") +
    labs(x = "Compound Hydrophobicity Score", y = "Density", title = "Distribution of Compound Hydrophobicity Scores") +
    theme_bw()

#ggplot(compound_df, aes(x = compound_hydropathy_score)) +
#    geom_density(aes(y = after_stat(density)), fill = "blue", alpha = 0.2) +
#    stat_function(fun = dnorm, args = list(mean = mixed_model$mu[1], sd = mixed_model$sigma[1]), colour = "red") +
#    stat_function(fun = dnorm, args = list(mean = mixed_model$mu[2], sd = mixed_model$sigma[2]), colour = "green") +
#    labs(x = "Compound Hydrophobicity Score", y = "Density", title = "Distribution of Compound Hydrophobicity Scores") +
#    theme_bw()
```